Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2018-11-01 anzjs2018/sections/03_models.tex
Clean up dir

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance travelled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are included in the state and estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We use a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modelling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function, 
discussed in section~\ref{sec:pf_update}.
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

\begin{figure}[p]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{
            Vehicle state is mapped to observation space by the
            measurement function $h$.   
        }
        % The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{
            The transition function $f$ predicts the future state 
            of each particle.
        }
            % The transition function $f$ models the behaviour of each particle,
            % including changes in acceleration and stopping behaviour at bus stops,
            % which can result in multiple modes.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    % \begin{subfigure}[t]{0.48\textwidth}
    %     \centering
    %     \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
    %     \caption{Each particle's distance from the true observation (red cross) 
    %         is calculated using
    %         the geographic distance between two coordinates, which can then be used
    %         to calculate it's weight.}
    %     \label{fig:pf_state_update}
    % \end{subfigure}\;\;
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{
            The state is updated by weighted resampling based on distance
            of particle to observation.
        }
        % Resampling occurs using likelihood weights that use the distance
        %     between the particle's location and the vehicles reported location (red cross),
        %     resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{
        The particle filter approximates vehicle state using a set of 
        discrete points (a), which are each mutated independently (b)
        to predict the next state.
        The measurement function $h$ allows state particle state
        to be mapped to a GPS coordinate (a),
        which can be compared to the observed location (c)
        to calculate each particles likelihood and perform resampling.
    }
        % The vehicles state is estimated using a set of discrete points, or particles,
        % each of which is independently mutated to make predictions of the future state.
        % Multimodality is easily handled, as demonstrated here,
        % without degeneration of the filter.}
    \label{fig:pf_state}
\end{figure}

@@

In a particle filter, the posterior distribution of the state at time $t_{k-1}$
is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
each with an associated weight $W_{k-1}^{(i)}$
\begin{equation}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
:= \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation}
using parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, 
the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
(Section~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
\begin{equation*}
W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
}
\end{equation*}
generating a posterior estimate of the state
\begin{equation}
p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
\left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
\end{equation}

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size 
\begin{equation}
\label{eq:neff}
N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
\end{equation}
falls below a specified threshold $N_{\text{thres}}$
\citep{}.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialised to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's speed changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on the values used by \cite{Hans_2015};
future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.

The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
The $stop\_index$ function returns the index of a stop,
given a distance travelled and set of stop distances.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
which involves changing the vehicle's speed each second (i.e., acceleration).
Lines~13--25 handle arrival behaviour at stops:
if the particle passes the next stop,
it has a chance of stopping (lines~18--19),
in which case the dwell time is sampled and subtracted from the 
time remaining to travel (lines~20--21).
Finally, to track travel time along segments,
at each point where the vehicle travels (lines 9 and 26),
the travel time for the current segment, $z_j$, is incremented.
The mutation ends once the travel time $\delta$ reaches zero,
or when the vehicle reaches the end of the route (lines 10, 14).
Once complete, the particle's new state is returned (line 29).

\renewcommand{\algorithmicrequire}{\textbf{Start:}}
\newcommand{\algorithmicbreak}{\textbf{break}}
\newcommand{\Break}{\State \algorithmicbreak}
\begin{algorithm}[tb]
    \caption{Particle mutation function.}
    \label{fig:algorithm}
    \begin{algorithmic}[2]
    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
    Initialisation $\delta \gets t_k - t_{k-1}$, 
        $j \gets stop\_index(x, \boldsymbol{S})$
    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
        \Comment{Remain longer at stop}
        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
        \State $\delta \gets \delta - \gamma - d_j$
    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
        \Comment{Occasionally bus stopped due to traffic}
        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
        \State $\delta \gets \delta - w$
        \State $z_j \gets z_j + w$
        \Comment{Congestion counts as travel time}
    \EndIf
    \While {$\delta > 0$ and $x < S_J$} 
        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
        \Comment{System noise, truncated to be non-negative}
        \State $x \leftarrow x + \dot x$
        \If {$x \geq S_{j+1}$}
            \If {$j+1 = J$}
                \Comment{Break at end of route}
                \State $x \gets S_J$
                \Break
            \EndIf
            \State $u_j \sim \mathrm{U}(0,1)$
            \Comment{Simulate vehicle stopping}
            \If {$u_j < \pi$}
                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
                \State $\delta \gets \delta - \gamma - d_j$
                \State $x \gets S_j$
            \EndIf
            \State $j\gets j+1$
        \EndIf
        \State $\delta \gets \delta - 1$
        \State $z_j \gets z_j + 1$
        \Comment{Track travel time along road}
    \EndWhile
    \\
    \Return $x, \dot x, \boldsymbol z$
    \end{algorithmic}
\end{algorithm}

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivariate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
The geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is calculated using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discrete nature of the particle filter state estimate;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance travelled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

@@

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$,
and is assumed to be constant over time 
with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
using the uncertainty of particle estimates.
The model from (\ref{eq:rbe_model}) now reduces to
\begin{equation}
\begin{split}
\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation}
where $\Delta_c = t_c - t_{c-1}$.

Since multiple vehicles can travel along a road simultaneously,
we have used an information filter to implement the network model.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state
in a single iteration.

>>>>

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance travelled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are included in the state and estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector
$\bX_k = \left[x_k\ \dot x_k\ \bz\right]^\top$.
% \begin{equation*}
% \label{eq:vehicle_state}
% \bX_k = 
% \begin{bmatrix}
%     x_k \\ \dot x_k \\ \bz
% \end{bmatrix}.
% \end{equation*}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$
% \begin{equation}
% \label{eq:vehicle_obs}
% \bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
% \end{equation}

We use a particle filter to estimate $\bX_k$,
due to its high flexibility and use in recent 
transit vehicle modelling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function, 
discussed in section~\ref{sec:pf_update}.
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

\begin{figure}[p]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{
            Vehicle state is mapped to observation space by the
            measurement function $h$.   
        }
        % The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{
            The transition function $f$ predicts the future state 
            of each particle.
        }
            % The transition function $f$ models the behaviour of each particle,
            % including changes in acceleration and stopping behaviour at bus stops,
            % which can result in multiple modes.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    % \begin{subfigure}[t]{0.48\textwidth}
    %     \centering
    %     \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
    %     \caption{Each particle's distance from the true observation (red cross) 
    %         is calculated using
    %         the geographic distance between two coordinates, which can then be used
    %         to calculate it's weight.}
    %     \label{fig:pf_state_update}
    % \end{subfigure}\;\;
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{
            The state is updated by weighted resampling based on distance
            of particle to observation.
        }
        % Resampling occurs using likelihood weights that use the distance
        %     between the particle's location and the vehicles reported location (red cross),
        %     resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{
        The particle filter approximates vehicle state using a set of 
        discrete points (a), which are each mutated independently (b)
        to predict the next state.
        The measurement function $h$ allows each particle state
        to be mapped to a GPS coordinate (a),
        which can be compared to the observed location (c)
        to calculate each particle's likelihood to compute resampling weights.
    }
        % The vehicles state is estimated using a set of discrete points, or particles,
        % each of which is independently mutated to make predictions of the future state.
        % Multimodality is easily handled, as demonstrated here,
        % without degeneration of the filter.}
    \label{fig:pf_state}
\end{figure}

@@

In a particle filter, the posterior distribution of the state at time $t_{k-1}$
is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
each with an associated weight $W_{k-1}^{(i)}$
\begin{equation*}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
:= \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation*}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation*}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation*}
using parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, 
the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
(Section~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
\begin{equation*}
W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
}
\end{equation*}
generating a posterior estimate of the state
\begin{equation*}
p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
\left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
\end{equation*}

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
falls below a specified threshold $N_{\text{thres}}$.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialised to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's speed changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on those used by \cite{Hans_2015};
future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.

The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
the $stop\_index$ function returns the index of a stop,
given a distance travelled and set of stop distances $\boldsymbol{S}$.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
which involves changing the vehicle's speed each second (i.e., acceleration).
Lines~13--25 handle arrival behaviour at stops:
if the particle passes the next stop,
it has a chance of stopping (lines~18--19),
in which case the dwell time is sampled and subtracted from the 
time remaining to travel (lines~20--21).
Finally, to track travel time along segments,
at each point where the vehicle travels (lines 7 and 26),
the travel time for the current segment, $z_j$, is incremented.
The mutation ends once the travel time $\delta$ reaches zero,
or when the vehicle reaches the end of the route (lines 10, 14).
Once complete, the particle's new state is returned (line 29).

\renewcommand{\algorithmicrequire}{\textbf{Start:}}
\newcommand{\algorithmicbreak}{\textbf{break}}
\newcommand{\Break}{\State \algorithmicbreak}
\begin{algorithm}[t]
    \caption{Particle mutation function.}
    \label{fig:algorithm}
    \begin{algorithmic}[2]
    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
    Initialisation $\delta \gets t_k - t_{k-1}$, 
        $j \gets stop\_index(x, \boldsymbol{S})$
    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
        \Comment{Remain longer at stop}
        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
        \State $\delta \gets \delta - \gamma - d_j$
    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
        \Comment{Occasionally bus stopped due to traffic}
        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
        \State $\delta \gets \delta - w$
        \State $z_j \gets z_j + w$
        \Comment{Congestion counts as travel time}
    \EndIf
    \While {$\delta > 0$ and $x < S_J$} 
        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
        \Comment{System noise, truncated to be non-negative}
        \State $x \leftarrow x + \dot x$
        \If {$x \geq S_{j+1}$}
            \If {$j+1 = J$}
                \Comment{Break at end of route}
                \State $x \gets S_J$
                \Break
            \EndIf
            \State $u_j \sim \mathrm{U}(0,1)$
            \Comment{Simulate vehicle stopping}
            \If {$u_j < \pi$}
                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
                \State $\delta \gets \delta - \gamma - d_j$
                \State $x \gets S_j$
            \EndIf
            \State $j\gets j+1$
        \EndIf
        \State $\delta \gets \delta - 1$
        \State $z_j \gets z_j + 1$
        \Comment{Track travel time along road}
    \EndWhile
    \\
    \Return $x, \dot x, \boldsymbol z$
    \end{algorithmic}
\end{algorithm}

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivariate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation*}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation*}
The geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation*}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation*}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging gives
\begin{equation*}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right).
\end{equation*}

The likelihood of the data given a particle's state estimate 
can now be calculated as
\begin{equation*}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation*}
It is worth noting that this representation of the likelihood is only
possible due to the discrete nature of the particle filter state estimate;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance travelled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

@@

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$,
and is assumed to be constant over time 
with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
using the uncertainty of particle estimates.
The model from (\ref{eq:rbe_model}) reduces to
\begin{equation*}
\begin{split}
\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation*}
where $\Delta_c = t_c - t_{c-1}$.

Since multiple vehicles can travel along a road simultaneously,
we used an information filter to implement the network model.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state
in a single iteration.
